Non-linear Models
=================

```{r}
require(ISLR)
attach(Wage)
```

Polynomials
-----------
Single variable: age
```{r}
fit <- lm(wage~poly(age,4), data = Wage)
summary(fit)
```

The function poly generates a basis of *orthogonal polynomials*.

The parament raw=TRUE can be added to return the polynomials directly:

```{r}
fit <- lm(wage~poly(age,4, raw = TRUE), data = Wage)
summary(fit)
```

To make a plot of the fitted function along with the standard errors:
```{r, fig.width=7, fig.height=6}
age.lims <- round(range(age),0)
age.grid <- seq(from = age.lims[1], to = age.lims[2])
preds <- predict(fit, newdata = list(age = age.grid), se = TRUE)
se.bands <- cbind(preds$fit - preds$se.fit * 1.96
                  , preds$fit + preds$se.fit * 1.96)
plot(wage~age, col = "darkgrey")
lines(age.grid, preds$fit, lwd=2, col = "blue")
matlines(age.grid, se.bands, lty = 2, col = "blue")
```

ANOVA to determine polynomial degree
====================================
Using a hypothesis test to determine the degree of polynomial to use.

```{r}
fit.1=lm(wage∼age,data=Wage)
fit.2=lm(wage∼poly(age,2),data=Wage)
fit.3=lm(wage∼poly(age,3),data=Wage)
fit.4=lm(wage∼poly(age,4),data=Wage)
fit.5=lm(wage∼poly(age,5),data=Wage)
anova(fit.1,fit.2,fit.3,fit.4,fit.5)
```

Polynomial Logistic Regression
------------------------------
```{r}
fit <- glm(I(wage>250)~poly(age,4), data = Wage, family = binomial)
summary(fit)
preds <- predict(fit, newdata = list(age = age.grid), se = TRUE)
se.bands <- preds$fit + cbind(fit = 0
                              , lower = -1.96 * preds$se.fit
                              , upper = 1.96 * preds$se.fit)
```

This computed on the logit scale. We can convert to the probability scale by mapping to the inverse logit:

$$p=\frac{e^\eta}{1+e^\eta}$$

```{r}
prob.bands <- exp(se.bands)/(1+exp(se.bands))
matplot(age.grid, prob.bands, col = "blue", lty = c(1,2,2), type = "l", ylim = c(0, 0.1))
points(jitter(age), I(wage >250)/10, pch="i", cex = 0.5)
```

Splines
-------
Splines are more lexible that polynoms
```{r}
require(splines)
fit <- lm(wage~bs(age, knots = c(25,40,60)), data = Wage)
plot(age,wage, col = "darkgrey")
lines(age.grid, predict(fit, list(age = age.grid)), col = "darkgreen", lwd = 2)
abline(v=c(25,40,60), lty=2, col = "darkgreen")
```
bs by default gives a cubic polynom by default. They are only discontinuous in the third derivative (for cubic).

Smoothing splines don't require a knot selection.
```{r, eval = FALSE}
fit <- smooth.spline(age, wage, df=16)
lines(fit, col = "red", lwd = 2)
```

Or can tell it to use cross validation to choose df automatically
```{r, eval = FALSE}
fit <- smooth.spline(age, wage, cv = TRUE)
lines(fit, col = "purple", lwd = 2)
fit
```

