```{r prologue, results='hide', echo=FALSE}
knitr::opts_chunk$set(warning = FALSE
                      , message = FALSE
                      , echo = FALSE
                      )
```

```{r setup}
require(vcd)
require(vcdExtra)
require(MASS)
require(lmtest)
require(effects)
require(car)
require(rms)
require(VGAM)
require(reshape2)
require(ggplot2)
require(directlabels)
require(nnet)
```

---
title: "VCD Exercise 8"
author: "Julian Hatwell"
date: "March 28, 2016"
output: html_document
---

## 8.1 
For the women's labor force participation data (Womenlf), the response variable, partic, can be treated as ordinal by using

Use the methods in Section 8.1 to test whether the proportional odds model holds for these data.

```{r}
data("Womenlf", package = "car")
Womenlf$partic <- ordered(Womenlf$partic, levels = c("not.work", "parttime", "fulltime"))
str(Womenlf)
```

```{r}
# need to compare PO with non-PO models (latter allows slope to change for different coeffs)
wlf.po <- vglm(partic ~ hincome + children, data = Womenlf,
                family = cumulative(parallel = TRUE))
wlf.po

# parrallel = FALSE is the NPO model
wlf.npo <- vglm(partic ~ hincome + children, data = Womenlf,
                 family = cumulative(parallel = FALSE))
wlf.npo

# this coef function is specific to the vglm object
coef(wlf.po, matrix = TRUE) # only the intercept changes
coef(wlf.npo, matrix = TRUE) # all the coefs change
```

*These coefs are so wildly different it's difficult to be sure that there isn't a mistake in the calcs.*

```{r}
VGAM::lrtest(wlf.npo, wlf.po) # equivalent to anova for vglm objects
```

*Is Chisq so vanishingly small that it's not even printed?*

```{r}
# rms plot
op <- par(mfrow=c(1,2))
plot.xmean.ordinaly(partic ~ hincome + children,
                  data = Womenlf,
                    lwd=2, pch=16, subn=FALSE)
par(op)
```

*This graphs show support for the case against the PO model. It doesn't hold for hincome and children is not well differentiated between not.work and parttime.*

## 8.2 
The data set housing in the MASS package gives a 3 * 3 * 4 * 2 table in frequency form relating (a) satisfaction (Sat) of residents with their housing (High, Medium, Low), (b) perceived degree of influence (Infl) they have on the management of the property (High, Medium, Low), (c) Type of rental (Tower, Atrium, Apartment, Terrace), and (d) contact (Cont) residents have with other residents (Low, High). Consider satisfaction as the ordinal response variable.

```{r}
data("housing", package = "MASS")
housing <- within(housing, { 
  Infl <- ordered(Infl, levels = c("Low", "Medium", "High"))
  Cont <- ordered(Cont, levels = c("Low", "High"))
  Type <- ordered(Type, levels = c("Tower", "Atrium", "Apartment", "Terrace"))
})
str(housing)
```

(a)	Fit the proportional odds model with additive (main) effects of housing type, influence in management, and contact with neighbors to this data. (Hint: Using polr(), with the data in frequency form, you need to use the weights argument to supply the Freq variable.)

```{r}
hou.polr <- polr(Sat ~ Infl + Cont + Type, weights = Freq
                  , data = housing, Hess = TRUE)
summary(hou.polr)
Anova(hou.polr)
```

(b)	Investigate whether any of the two-factor interactions among Infl, Type, and Cont add substantially to goodness of fit of this model. (Hint: use stepAIC(), with the scope formula ~.^2 and direction="forward")

```{r}
hou.step <- stepAIC(hou.polr, scope = ~.^2, direction = "forward", trace = FALSE)
Anova(hou.step)
hou.step$anova
```

(c)	For your chosen model from the previous step, use the methods of Section 8.1.5 to plot the probabilities of the categories of satisfaction.

```{r}
hou.chosen <- polr(Sat ~ Infl + Cont + Type + Infl:Type + Cont:Type
                   , weights = Freq
                   , data = housing, Hess = TRUE)

hou.fitp <- cbind(housing,
                   predict(hou.chosen, type = "probs"))
head(hou.fitp)

# this is an unpivot operation
plotdata <- melt(hou.fitp
                 , id.vars = c("Infl", "Type", "Cont")
                 , measure.vars = c("Low", "Medium", "High"),
                variable.name = "Sat",
                value.name = "Probability")

gg <- ggplot(plotdata, aes(x = as.numeric(Type), y = Probability, colour = Sat)) +
  geom_line(size = 2.5) + 
  theme_bw() + #xlim(0.5, 4.5) +
  geom_point(color = "black", size = 1.5) +
  facet_grid(Infl~Cont, labeller = label_both) +
  xlab("Type") + scale_x_discrete(limits=c("Tower", "Atrium", "Apartment", "Terrace"), expand = c(0.1, 0.1))
  # scale_x_manual(levels = c("Tower", "Atrium", "Apartment", "Terrace"))
direct.label(gg)
```

(d)	Write a brief summary of these analyses, interpreting how satisfaction with housing depends on the predictor variables. 

*It's a rather complext inter-relationship of the variables. Generally, satisfaction is highest with influence is highest. Tower accommodation generally gives higher satisfaction. Contact with neighbours improves satisfaction. Having more influence and/or more contact changes satisfaction with the specific housing types. For example, people in terraces and apartments don't like much contact with neighbours.*

## 8.3 
The data TV on television viewing was analyzed using correspondence analysis in Example 6.4, ignoring the variable Time, and extended in Exercise 6.9. Treating Network as a three-level response variable, fit a generalized logit model (Section 8.3) to explain the variation in viewing in relation to Day and Time. The TV data is a three-way table, so you will need to convert it to a frequency data frame first.

```{r}
data("TV", package="vcdExtra")
TV.df <- data.frame(structable(~Time + Day + Network, TV))
TV.df <- within(TV.df, {
  Time <- as.numeric(Time) * 0.25 + 7.75
  Day <- ordered(Day, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
})
```

(a)	Fit the main-effects model, Network ~ Day + Time, with multinom(). Note that you will have to supply the weights argument because each row of TV.df represents the number of viewers in the Freq variable.

```{r}
TV.mn <- multinom(Network ~ Day + Time, weights = Freq, data = TV.df, Hess = TRUE)
Anova(TV.mn)
stats <- summary(TV.mn, Wald = TRUE)
z <- stats$Wald.ratios
p <- 2 * (1 - pnorm(abs(z)))
zapsmall(p)
```

(b)	Prepare an effects plot for the fitted probabilities in this model.

```{r}
plot(Effect(c("Time", "Day"), TV.mn)
     , style = "stacked"
     , key.args = list(x = .05, y = .9))
```

(c)	Interpret these results in comparison to the correspondence analysis in Example 6.4.

*The daily trends agree with the correspondence analysis. However, the correspondence analysis was collapsed over time and so couldn't reveal any information about trends over the course of the evening. The effect plot shows a clear trend each day of more people watching CBS in the early part of the evening and then switching to other channels as the evening progresses.*

*Just for fun, a model plot:*

```{r}
TV.pred <- cbind(TV.df, predict(TV.mn), predict(TV.mn, type = "probs"))
some(TV.pred)
```

*Reshaping*

```{r}
TV.melt <- melt(TV.pred
                , id.vars = c("Time", "Day", "Network")
                , measure.vars = c("ABC", "CBS", "NBC")
                , variable.name = "Viewing.pred"
                , value.name = "Probability")
some(TV.melt)
```

```{r}
gg <- ggplot(TV.melt,
             aes(x = Time, y = Probability, colour = Viewing.pred)) + 
  facet_grid(~ Day, 
             labeller = label_both) + 
  geom_line(size = 2) +
  theme_bw() +
  scale_x_continuous(limits = c(8, 11))
  #scale_y_continuous(limits = c(0, 0.9))   
  direct.label(gg, list("last.qp", dl.trans(y = y + 0.1)))
```

## 8.4 
Refer to Exercise 5.10 for a description of the Vietnam data set in vcdExtra. The goal here is to fit models for the polytomous response varialble in relation to year and sex.

```{r}
data("Vietnam", package = "vcdExtra")
viet <- within(Vietnam, {
  year <- ordered(year)
  response <- ordered(response)
})
```

(a)	Fit the proportional odds model to these data, allowing an interaction of year and sex.

```{r}
# need to compare PO with non-PO models (latter allows slope to change for different coeffs)
viet.po <- vglm(response ~ year * sex
                , weights = Freq
                , data = viet
                , family = cumulative(parallel = TRUE))

viet.po

# parrallel = FALSE is the NPO model
viet.npo <- vglm(response ~ year * sex
                , weights = Freq
                , data = viet
                , family = cumulative(parallel = FALSE))
viet.npo

# this coef function is specific to the vglm object
coef(viet.po, matrix = TRUE) # only the intercept changes
coef(viet.npo, matrix = TRUE) # all the coefs change
```

(b)	Is there evidence that the proportional odds assumption does not hold for this data set? Use the methods described in Section 8.1 to assess this.

```{r}
VGAM::lrtest(viet.npo, viet.po) # equivalent to anova for vglm objects
```

```{r}
# rms plot
viet.xpd <- expand.dft(viet)
viet.xpd <- within(viet.xpd, {
    year <- ordered(year)
  response <- ordered(response)
})
plot.xmean.ordinaly(response ~ year * sex
                    , data = viet.xpd
                    , lwd=2, pch=16
                    , subn=FALSE)
```

*Very good support for the model apart from this divergence for males in year 5.*

(c)	Fit the multinomial logistic model, also allowing an interaction. Use car::Anova () to assess the model terms.

```{r}
viet.mn <- multinom(response ~ year * sex
                    , weights = Freq
                    , data = viet
                    , Hess = TRUE)
Anova(viet.mn)
stats <- summary(viet.mn, Wald = TRUE)
z <- stats$Wald.ratios
p <- 2 * (1 - pnorm(abs(z)))
zapsmall(p)
```

*The interaction term is barely significant*

(d)	Produce an effect plot for this model and describe the nature of the interaction.

```{r}
plot(Effect(c("sex", "year"), viet.mn),
     style = "stacked", key.args = list(x = .05, y = .9))
```

*There is not a terribly distinguishable pattern for females, other than perhaps a higher number of A and B in year 1. Males on the other hand appear to have a strong trend away from A towards more in C and D with increasing year.*

(e)	Fit the simpler multinomial model in which there is no effect of year for females and the effect of year is linear for males (on the logit scale). Test whether this model is significantly worse than the general multinomial model with interaction.

```{r}
viet <- within(viet, {
  yearMale <- ifelse(sex=="Male", year, 0)
})
viet.mn2 <- multinom(response ~ sex + yearMale
                    , weights = Freq
                    , data = viet
                    , Hess = TRUE)
Anova(viet.mn2)
anova(viet.mn2, viet.mn)
stats <- summary(viet.mn2, Wald = TRUE)
z <- stats$Wald.ratios
p <- 2 * (1 - pnorm(abs(z)))
zapsmall(p)

plot(Effect(c("sex", "yearMale"), viet.mn2),
     style = "stacked", key.args = list(x = .05, y = .9))
```